---
title: "StatR 502 Homework 1"
author: "Rebecca Hadi"
date: "Due Thursday, Jan. 11, 2018 at 6:30 pm"
output:
  pdf_document:
    toc: yes
---

## 1: Overplotting

Load the data for this problem (the file `hw1data.rdata`) using the `load()` command. This will create a `data.frame` in your workspace called `pr1` (no assignment with `<-` or `=` needed!). The data frame has two columns, `x` and `y`, and there's a surprise hidden in it. Your job is to find the surprise through some exploratory plots. Once you've found it, make a plot that shows it nicely, and present that as your solution.

This problem highlights one of the main drawbacks of `ggplot2`: it can be RAM-intensive and slow with large data sets. Usually, however, you don't need half a million points in a single plot. Transparency can help a lot, but
subsets and statistical summaries can do a nice job without taxing your computer or your patience.


```{r,message=FALSE, warning=FALSE}
#load data
load("C:\\Users\\BT0x\\Downloads\\hw1data.rdata")

#bring in ggplot2
library(ggplot2)
#build plot
ggplot(data = pr1, aes(x,y)) + 
  geom_point(alpha = .05) + 
  theme_light() + 
  ggtitle("Smile!")  

```



## 2: Exploring new options

Make three ggplots exploring different options. At least one of the plots should use facets, and none of them should be plain scatterplots (if you use `geom_point`, complement it with another `geom` or `stat`). You can choose any dataset(s) you've worked with in StatR 501, from other problems on this homework, or from your work/interests.

Pick one of your plots to polish, and spend an extra 10-15 minutes on it adding nice labels, adjusting a theme element or two, making sure any factors are ordered in a meaningful way, etc.

Some suggestions for geoms (but feel free to explore further!): `geom_rug`, `geom_boxplot`, `geom_text`, `geom_violin`, `stat_smooth`.

* *First, I need to load and prepare my data for plotting. I decided to download a dataset from Kaggle that contained 17 years of electronic music reviews from the website "Resident Advisor"*

```{r, message=FALSE, warning=FALSE}
#load & look at data 
music <- read.csv("Hw1 Data\\RA_cleaned.csv")

#reorder factor level of release month for later plotting
music$release_month <- factor(music$release_month, levels = c("January", "February", "March","April","May","June","July","August","September","October","November","December"))

```

Plot 1
```{r}
ggplot(music,aes(x = release_year, y = num_comments, col = rating, shape = release_type)) +
    geom_point(alpha = 1) + 
    theme_classic() +  
    coord_cartesian(ylim = c(0,100)) + 
    ggtitle("Number of Comments over time by Rating and Release Type") + 
    labs(x = "Album Release Year", 
         y =  "Count of Comments", 
         color = "Rating", 
         shape = "Release Type")
```

Plot 2 
```{r}
ggplot(music,aes(x = release_year)) +
    geom_bar(fill = "steelblue") + 
    facet_wrap (~release_type) + 
    theme_classic() + 
    ggtitle("Count of Releases Reviewed by Release Type and Year") + 
    labs(x = "Album Release Year", 
         y =  "Count of Releases Reviewed")
```

Plot 3 
```{r}
ggplot(music, aes(x = release_month, col = release_type)) +  
     stat_summary_bin(aes(y = rating), fun.y = "mean", geom = "point") + 
     theme_classic() + 
     ggtitle("Average Release Rating by Month and Release Type") + 
     labs(x = "Release Month", 
          y = "Average Rating", 
          color = "Release Type")
```


## Problems from Gelman & Hill

Section 3.9 (pp. 49-51), **do problems 2, 3, and 5**. In your write-up, please label them as G&H 2, G&H 3, and G&H 5. The `se.coef()` function in G&H 3 is part of the `arm` package (written to accompany the book).

## G&H 2

Suppose that, for a certain population, we can predict log earnings from log
height as follows:

A person who is 66 inches tall is predicted to have earnings of $30,000.
 Every increase of 1% in height corresponds to a predicted increase of 0.8% in
earnings.

(a) The earnings of approximately 95% of people fall within a factor of 1.1 of
predicted values.
Give the equation of the regression line and the residual standard deviation
of the regression. 
(b) Suppose the standard deviation of log heights is 5% in this population. What,
then, is the R2 of the regression model described here?


**For G&H 2:** The logs add a little twist to this problem. We'll be talking about transformations - especially log transformations - next week. A couple clarifications/hints:

- "1% change in $x$ results in 0.8% change in $y$" means that the slope of $\log y$ versus $\log x$ is 0.8.

- "Fall within a factor of 1.1" on the untransformed scale means "plus or minus 0.1" on the log scale. (Well, it really means $\pm \log(1.1)$ but $\log(1.1) = `r log(1.1)`$ so we'll call 0.1 a good-enough approximation.)



## G&H 3



## G&H 5



**For G&H 5:** the data can be found in the `AER` package, it's called `TeachingRatings`. The column names are different from those called out in the book, see `?TeachingRatings` for details. In part (b), let's consider "some other models" to mean "two or three" other models that you explain.

